Correctly Optimizing Abstract Abstract Machines

### 2013-03-29 paper/paper.tex
We actually do better than 25 times slower now. Hooray.

The only analyzers we were able to find that proved capable of
analyzing the full suite of benchmarks considered here were the Soft
Typing system of Wright and
Cartwright~\cite{dvanhorn:Wright1997Practical} and, in many ways its
successor, the Polymorphic splitting system of Wright and
Jagannathan~\cite{dvanhorn:wright-jagannathan-toplas98}.\footnote{This
  is not a coincidence; these papers set a high standard for
  evaluation, which we consciously aimed to approach.}  Unfortunately,
these analyses compute an inherently different and incomparable form
of analysis.  Consequently, we have omitted a complete comparison with
these implementations.  The AAM approach provides more precision in
terms of temporal-ordering of program states, which comes at a cost
that can be avoided in constraint-based approaches.  Consequently
implementation techniques cannot be ``ported'' between these two
approaches.  However, our optimized implementation is within an order
of magnitude of the performance of Wright and Jaganathan's analyzer.
Although we would like to improve this to be more competitive, the
optimized AAM approach still has many strengths to recommend it in
terms of precision, ease of implementation and verification, and rapid
design. We can get closer to their performance by relying on the
representation of addresses and the behavior of $\alloc$ to
pre-allocate most data structures and split the abstract store out
into parts that are more quickly accessed and updated. The first two
of our optimizations can still be applied to an analysis that does
abstract garbage collection~\cite{dvanhorn:Might:2006:GammaCFA},
whereas the polymorphic splitting implementation is tied strongly to a
single-threaded store.

@@

This effort increases substantially as the complexity of the analyzed
language increases.  Both the work of maintaining the concrete
semantics and constraint system (and the relations between them) must
be scaled simultaneously.  However, constraint systems, which have
been extensively studied in their own right, enjoy efficient
implementation techniques and can be expressed in declarative logic
languages that are heavily
optimized~\cite{dvanhorn:bravenboer-smaragdakis-oopsla09}.
Consequently, constraint-based analyses can be computed quickly.  For
example, Jagannathan and Wright's polymorphic splitting
implementation~\cite{dvanhorn:wright-jagannathan-toplas98} analyses
the \Church{} benchmark about 25 times faster than the fastest
implementation considered here.  These analyses compute very different
things, so the performance comparison is not apples-to-apples.

>>>>

The only analyzers we were able to find that proved capable of
analyzing the full suite of benchmarks considered here were the Soft
Typing system of Wright and
Cartwright~\cite{dvanhorn:Wright1997Practical} and, in many ways its
successor, the Polymorphic splitting system of Wright and
Jagannathan~\cite{dvanhorn:wright-jagannathan-toplas98}.\footnote{This
  is not a coincidence; these papers set a high standard for
  evaluation, which we consciously aimed to approach.}  Unfortunately,
these analyses compute an inherently different and incomparable form
of analysis.  Consequently, we have omitted a complete comparison with
these implementations.  The AAM approach provides more precision in
terms of temporal-ordering of program states, which comes at a cost
that can be avoided in constraint-based approaches.  Consequently
implementation techniques cannot be ``ported'' between these two
approaches.  However, our optimized implementation is within an order
of magnitude of the performance of Wright and Jaganathan's analyzer.
Although we would like to improve this to be more competitive, the
optimized AAM approach still has many strengths to recommend it in
terms of precision, ease of implementation and verification, and rapid
design. We can get closer to their performance by relying on the
representation of addresses and the behavior of $\alloc$ to
pre-allocate most data structures and split the abstract store out
into parts that are more quickly accessed and updated. Our semantic
optimizations can still be applied to an analysis that does abstract
garbage collection~\cite{dvanhorn:Might:2006:GammaCFA}, whereas the
polymorphic splitting implementation is tied strongly to a
single-threaded store.

@@

This effort increases substantially as the complexity of the analyzed
language increases.  Both the work of maintaining the concrete
semantics and constraint system (and the relations between them) must
be scaled simultaneously.  However, constraint systems, which have
been extensively studied in their own right, enjoy efficient
implementation techniques and can be expressed in declarative logic
languages that are heavily
optimized~\cite{dvanhorn:bravenboer-smaragdakis-oopsla09}.
Consequently, constraint-based analyses can be computed quickly.  For
example, Jagannathan and Wright's polymorphic splitting
implementation~\cite{dvanhorn:wright-jagannathan-toplas98} analyses
the \Church{} benchmark about 5.5 times faster than the fastest
implementation considered here.  These analyses compute very different
things, so the performance comparison is not apples-to-apples.
